---
title: "Getting Started with Relational Data"
author: "Bret Larget"
output: html_document
---

- Professor Jessi Kehe contributed to these lecture notes

### Setup details

- You will need the packages **tidyverse**, **lubridate**, **nasaweather** for this lecture. 

- Be sure to read R4DS Chapter 13 for information on relational data

- We are going to consider a few simple examples to help you get started with relational data.
- Practice on more complicated examples will come during the lectures and as the semester progresses.
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE,
                      autodep=TRUE, cache.comments=FALSE)
library(tidyverse)
library(lubridate)
library(nasaweather)
```

## Primary Key

- A *primary key* for a data frame are the variable or variables that uniquely identify an observation.

- Let's try to find the primary key in the `nasaweather::atmos` data.  We will begin by running `?nasaweather::atmos` for details on the data, which may provide some hint at a primary key.
- When using `count()` on a variable or collection of variables that are a primary key, all counts should be equal to one.

### Help for Atomosphere Data

```{r}
##
?nasaweather::atmos

atmos = nasaweather::atmos
class(atmos)
dim(atmos)
names(atmos)
```

### Seek a primary key

> Is `temp` a primary key to the `atmos` data set?

```{r}
## temp
atmos %>% 
  count(temp) %>% 
  filter(n > 1)
```

- Not a primary key: many cases have the same temperature

> Is the pair (`lat`, `lon`) a primary key to the `atmos` data set?

```{r}
## lat (latitude) and long (longitude)
atmos %>% 
  count(lat, long) %>% 
  filter(n > 1)

## Looks like each latitude longitude pair appears 72 times
atmos %>% 
  count(lat, long) %>% 
  filter(n != 72)
```

- The location variables, `lat` and `long` do not form a primary key.

> Is the pair (`year`, `month`) a primary key to the `atmos` data set?

```{r}
## year and month
atmos %>% 
  count(year, month) %>% 
  filter(n > 1)

atmos %>% 
  count(year, month) %>% 
  filter(n != 576)
```

- The time variable pair, `year` and `month` also do not form a primary key.
    - There are 576 measurements at each distinct time.
    
> Is the set (`year`, `month`, `lat`, `lon`) a primary key to the `atmos` data set?

```{r}
## Look at time and location together
atmos %>% 
  count(lat, long, year, month) %>% 
  filter(n > 1)
```

- There is only a single observation at each location and time

## Mutating joins

- We were introduced to joins earlier during our introduction to dplyr,
but we will consider another example here.

- Consider the following two data frames. 

```{r drinks-discounts}
drinks = tibble(
  Drink = c("Milk Tea", "Espresso", "Latte", "Espresso"),
  Cafe = c("Happy Tea", "Daily Grind", "Daily Grind", "Bev's Beverages"),
  Price = c(4, 2, 3, 2)
)

drinks

discounts = tibble(
  Coupon = c("25%", "50%", "5%"),
  Location = c("Happy Tea", "Daily Grind", "The Roast")
)

discounts
```

- Note that the value in `Cafe` from data frame `drinks` and `Location` from data frame `discounts` have overlapping values and will be the basis of joining the two data sets together.
    - Often, the name or names of the columns on which to join are the same.
- Note that the syntax for joining by columns with different names is:
    - `by = c("name from data set 1" = "name from data set 2")
    - The quotes are needed as `c()` is a base R function so the **tidyverse** syntax that allows variable names to be written without the quotes does not hold within `c()`    

### Inner join

- An inner join keeps all cases present in both data sets.
- Note that we use the **tidyverse** idiom of `data set 1 %>% *_join(date set 2, ...)`
    - An alternative would be `*_join(data set 1, data set 2, ...)`
    
```{r}
drinks %>% 
  inner_join(discounts, by = c("Cafe" = "Location"))
```

- Notice that when two different column names are used to match for joining, the name of the first data frame is retained.

### Full join

- A full join keeps all cases from each data frame

```{r}
drinks %>% 
  full_join(discounts, by = c("Cafe" = "Location"))


drinks %>% 
  full_join(discounts, by = c("Cafe" = "Location"), keep=TRUE)
```

- Missing values are added where data is missing from one data frame or the other
- Notice that the argument `keep = TRUE` retains both joining variables

### Left join

- A left join keeps all cases from the first data frame and adds new columns from the second

    
```{r}
drinks %>% 
  left_join(discounts, by = c("Cafe" = "Location"))
```

- Typically, a left join retains the number of rows from the first data frame
    - More rows could be added if there were multiple matches from the second data frame to a single row of the first.

### Right join

- A right join keeps all rows from the second data frame and adds columns from the first.

```{r}
drinks %>% 
 right_join(discounts, by = c("Cafe" = "Location"))
```

## Filtering joins

- Filtering joins also look to match observations between the two data frame, but focus on the observations instead of the variables. 

- There are two filtering joins:  `semi_join()` and `anti_join()`.  

### Semi-join

- `semi_join()` returns all observations of `drinks` with a match in `discounts`
- this is a useful way to filter observations which match one of several values for a specific variable
- the columns in the returned table are the same as in the first data set

```{r filtering-joins}
drinks %>% 
  semi_join(discounts, by = c("Cafe" = "Location"))

discounts %>% 
  semi_join(drinks, by = c("Location" = "Cafe"))
```

### Anti joins

- `anti_join()` returns all observations of drinks that do *not* have a match in discounts
- this is another way to filter rows

```{r}
drinks %>% 
  anti_join(discounts, by = c("Cafe" = "Location"))

discounts %>% 
  anti_join(drinks, by = c("Location" = "Cafe"))
```

