---
title: "Getting Started with tidyr"
author: "Bret Larget"
output: html_document
---

- Professor Jesse Kehe made contributions to these lecture notes.

### Setup details

- You will need the packages **tidyverse**, **lubridate**, and **XML** for this lecture. 

- The following data files will be used and are assumed to be located two steps back from the working directory in a folder called `data/`.  Be sure to adjust the code if you have the data files in a different location.  
`"../../data/DEN-2018.html"`  
`"../../data/JFK-2018.html"`  
`"../../data/LAX-2018.html"`  
`"../../data/MSP-2018.html"`  
`"../../data/ORD-2018.html"`  
`"../../data/SFO-2018.html"` 

- Be sure to read R4DS Chapter 12 for information on **tidyr**.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE,
                      autodep=TRUE, cache.comments=FALSE)
library(tidyverse)
library(lubridate)
library(XML)
```

# tidyr 

- Tidy data satisfy the following criteria

    1. Each variable has its own column  
    2. Each observation has its own row
    3. Each value has its own cell
  
- Sometimes, we want to *reshape* data so that it is tidy (or tidy in a different way)
- For example, sometimes we read in data in a *wide* format where there are multiple columns with the same sort of data and the columns names contain information about another variable
    - An example of this is the airport waiting time data where the columns `all_n_0_15`, `all_n_16_30`, and so on each had a count of the number of passengers processed in some time interval and the column names had information about the limits of the time interval.
    - We could *reshape* this data so that:
        - all counts are in a single column
        - there is a new column with the time interval information
- The **tidyr** function to make this change is named `pivot_longer()`
- There is also a complementary `pivot_wider()` function.

## Airport Data

- Let's consider our airport timing data from last week.

```{r stuff-for-all-airports}
## Input:  x = HTML file
## Output: nicely formatted tibble
read_airport =  function(x)
{
  airport_column_names = c("airport",
                           "terminal",
                           "date",
                           "hour",
                           "us_avg_wait",
                           "us_max_wait",
                           "non_us_avg_wait",
                           "non_us_max_wait",
                           "all_avg_wait",
                           "all_max_wait",
                           "all_n_0_15",
                           "all_n_16_30",
                           "all_n_31_45",
                           "all_n_46_60",
                           "all_n_61_90",
                           "all_n_91_120",
                           "all_n_120_plus",
                           "all_excluded",
                           "all_total",
                           "all_flights",
                           "all_booths")

  airport = readHTMLTable(x)$`NULL` %>%
    slice(-(1:3))

  names(airport) = airport_column_names

  airport = airport %>%
    as_tibble() %>%
    mutate(across(ends_with("wait") | starts_with("all"), as.numeric)) %>% 
    mutate(across(where(is.factor),as.character)) %>% 
    mutate(date = mdy(date))
  
  return( airport )
}
```

### Read and Combine the Data

```{r read-data}
den = read_airport("../../data/DEN-2018.html")
jfk = read_airport("../../data/JFK-2018.html")
lax = read_airport("../../data/LAX-2018.html")
msp = read_airport("../../data/MSP-2018.html")
ord = read_airport("../../data/ORD-2018.html")
sfo = read_airport("../../data/SFO-2018.html")

awt = bind_rows(den,jfk,lax,msp, ord,sfo)
rm(den,jfk,lax,msp,ord,sfo)

str(awt)
```


### pivot_longer()

- Each observation in the data set is a single airport/terminal/date/hour.

- However, some of the columns contain data.
    - Namely, the `all_n_*` columns contain information in the names about the number of passengers who wait for given time intervals.
    
- An alternative way to represent the data would be to have all of these counts be in a single column.

- Then, an observation would entail a airport/terminal/date/hour/wait-interval combination.

- Let's do this for the subset of the data that are not the waiting times or booths.

```{r}
## eliminate some columns
awt_long = awt %>% 
  select(-contains("_wait"), -all_total, -all_flights, -all_booths)

## Show the column names
names(awt_long)

## Reshape the data by pivoting longer

awt_long = awt_long %>%
  pivot_longer(cols = starts_with("all_"),
               names_to = "wait_interval",
               values_to = "n")

names(awt_long)
```

- Examine the first row of `awt` and the first eight rows of the reshaped long-format data set.

```{r}
## First row of awt with selected columns
awt %>% 
  select(-contains("_wait"), -all_total, -all_flights, -all_booths) %>% 
  print(n = 1, width = Inf)

## First 8 rows of awt_long have data from first row of awt
awt_long %>% 
  print(n = 8)
```

### pivot_wider()

- The function `pivot_wider()` reshapes data by making it wider.
- A common usage is to create a wider table after doing a summary.
- Here, we demonstrate it's use after calculating the mean waiting time for all passengers by hour for each airport and then reshaping the data.

#### Create the summary table

- Average the mean waiting time for all passengers grouped by airport and hour

```{r}
awt_sum = awt %>% 
  group_by(airport, hour) %>% 
  summarize(mean_wait_time = mean(all_avg_wait))

head(awt_sum)
```

- Next, use `pivot_wider()` to reshape the data with a column for each airport and a row for each hour.
- The arguments are:
    - `names_from()` which determines which variable will produce the new column names
    - `values_from()` which determines which variable will produce the new values
- There will be missing data if an airport had no arrivals during a particular hour.

```{r}
## pivot_wider example
awt_sum_wide = awt_sum %>% 
  pivot_wider(names_from = airport, values_from = mean_wait_time)

awt_sum_wide %>% 
  print(n = Inf, width = Inf)
```

- The basic summary is most useful for further calculations
- The wider table is better to create a table to add to a report.

### Fill-in missing values

- Here, missing values are appropriate when there are no corresponding observations.
- However, at times, it makes sense to use a value such as zero.
- Let's do something similar, but calculate the total number of international arrivals per day for each airport at each hour.
- Here, missing means zero.

```{r}
## First total arrivals by airport and hour
awt_arrivals = awt %>% 
  group_by(airport, hour) %>% 
  summarize(flights = sum(all_flights))

dim(awt_arrivals)

awt_arrivals %>% 
  print(n = 30)
```

- Now make wider

```{r}
## default, with missing values
awt_arrivals_wide = awt_arrivals %>% 
  pivot_wider(names_from = airport,
              values_from = flights)

awt_arrivals_wide %>% 
  print(n = Inf, width = Inf)
```

- Do again, but replace `NA` with zero

```{r}
## default, with missing values
awt_arrivals_wide = awt_arrivals %>% 
  pivot_wider(names_from = airport,
              values_from = flights,
              values_fill = 0)

awt_arrivals_wide %>% 
  print(n = Inf, width = Inf)
```














